---
title: "About the Creation of this Package"
author: "Hauke Sonnenberg"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{about}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This package was created during the runtime of the KWB project 
[DSWT (Decentralised Treatment of Roadway Runoffs)](https://www.kompetenz-wasser.de/en/project/dswt-dezentrale-reinigung-von-strassenabfluessen/).
I want to document here what I did to the package since then.

## Convert Inline Documentation

In the package I already used so called inline documentation, i.e. I shortly
described functions and their arguments by means of special comments directly
in the source code. A function header, for example, looked like that:

```{r eval = FALSE}
readAndPlotAutoSamplerFiles <- function # readAndPlotAutoSamplerFiles
### readAndPlotAutoSamplerFiles
(
  filePaths,
  ### full path(s) to ORI Auto sampler log files PN_<yyyymmdd>_<station>.csv
  removePattern = "Power|Bluetooth|Modem|SMS|Sonde",
  ### regular expression pattern matching logged actions to be removed before
  ### plotting. Set to "" in order not to remove any action
  to.pdf = TRUE,
  ### if TRUE, graphical output is directed to PDF
  evtSepTime = 30*60
)
{
  # Here comes the body of the function
}
```

You could give a title after `function` (here just the name of the function,
`readAndPlotAutoSamplerFiles`, and a description direcly behind after three
comment characters, `###`. The example above reveals that I was too lazy to give
a good title and description. Instead, I just used the name of the function for
both. Then, again after `###` in each case and always directly following the
definition of an argument, you could document the meaning of the arguments of
the function. Again, the example above is not a good model as it lacks the
description of the last argument, `evtSepTime`.

I was then using the package "inlinedocs" to convert these comments into R
documentation files.

[Roxygen](https://cran.r-project.org/web/packages/roxygen2/vignettes/roxygen2.html)
is another system that allows to generate documentation from special comments in
the source code. As Roxygen has evolved to a quasi-standard for inline
documentation and is supported by the Development Environment
[RStudio](https://www.rstudio.com/), I converted all inlinedocs comments to
Roxygen comments. For that purpose I used the script `toRoxygen.R` that is under
Subversion control in the subfolder `RScripts/_OTHERS/`.

After conversion, the header of the function above looked (almost) like that:

```{r eval = FALSE}
#' Read and Plot Autosampler Files
#' 
#' @param filePaths full path(s) to ORI Auto sampler log files
#'   PN_<yyyymmdd>_<station>.csv
#' @param removePattern regular expression pattern matching logged actions to be
#'   removed before plotting. Set to "" in order not to remove any action
#' @param to.pdf if TRUE, graphical output is directed to PDF
#' 
readAndPlotAutoSamplerFiles <- function(
  filePaths,
  removePattern = "Power|Bluetooth|Modem|SMS|Sonde",
  to.pdf = TRUE,
  evtSepTime = 30 * 60
)
{
  # Here comes the body of the function
}
```

All documentation snippets are now altogether in a comment code block directly
above the function. In this block, all comments start with `#'` which indicates
that these are special comments that are interpreded by Roxygen. The block
starts with the title of the function. My converter script used my original
"title" that was just the name of the function but after conversion I put a new,
somehow meaningful title, "Read and Plot Autosampler Files". Roxygen supports
some keywords, so called tags, that start with the `@` character. In the above
example only the `@param` tag is used that indicates that the description of an
argument (parameter) of the function is following.

## Call Functions in Base Packages with their Package Name

When I ran "Check" on the package, I got some notes in case of calling functions
from the base packages "utils" or "graphics". A corresponding note reads like 
the following:

```
* checking R code for possible problems ... NOTE
plotAutoSamplerActions: no visible global function definition for
  ‘axis’
Undefined global functions or variables:
  axis
Consider adding
  importFrom("graphics", "axis")
to your NAMESPACE file.
```

I put `utils::` or `graphics::` in front of the function calls and the notes
disappeared.

## Export only the Top-Level Functions

It is good practice to organise your code in terms of functions. In order 

* to avoid code duplication, 
* to allow for a reuse of existing code and 
* to improve the readability of the code 

it is also good practice to split the code within functions up into further
sub-functions. This may result in a lot of functions and sub-functions. From a
developer's point of view you may say: the more functions the better, given the
fact that they have good interfaces and meaningful names. However, the user of
the package should not be confronted with the full bunch of functions. 
Therefore, only the top-level functions and not their sub-functions should be
"exported" from the package, i.e. made availalble to the user when calling
` library()`. In my original version, the package kwb.dswt did not respect this
recommendation and exported all functions by having the following line in the
`NAMESPACE` file:

```{r eval = FALSE}
exportPattern(".*")
```

I wanted to change this behaviour by using the Roxygen tag `@export` in the
Roxygen headers of the functions to be exported and to let Roxygen generate the
`NAMESPACE` file instead of manually maintaining this file. That arose the
following question: 

**What functions of the package are (or have ever been) used from a script outside of the package?**

In fact, this question and my thoughts about its answer motivated me to write
this vignette. I will (hopefully) propose a small script that finds out what
functions of a package are called from a set of given scripts. Here comes the
script that uses functions that I wrote in the scope of our project on research
data management, [FAKIN](https://www.kompetenz-wasser.de/en/project/fakin-research-data-management/).

You need to have the packages kwb.fakin and kwb.code installed. Install them from
GitHub with:

```{r eval = FALSE}
remotes::install_github("kwb-r/kwb.fakin")
remotes::install_github("kwb-r/kwb.code")
```


```{r include = FALSE}
usage <- read.csv(system.file(
  "extdata/function_usage.R", package = "kwb.dswt"
))
```

```{r eval = FALSE}
# Set the path to the root folder containing the scripts to analyse
root <- "~/HAUKE/R_Development/RScripts"

# Read all scripts below this root folder
tree <- kwb.code::parse_scripts(root = root)

# How many scripts were read?
length(tree)

# Check which functions from kwb.utils are used and how often
usage <- kwb.fakin::get_package_function_usage(
  tree, package = "kwb.dswt", simple = FALSE
)

# Show the table of function usage
usage
```
```{r eval = FALSE, include = FALSE}
write.csv(usage, "~/HAUKE/R/kwb.dswt/inst/extdata/function_usage.R")
```
